---
title: "Ingest CSV and SQL"
author: "John Little"
format: html
---

This time lets import a large CSV file. This example comes from a duckdb.org blog post about analyzing New Zeland census data. The data are a zipped collection of CSV files that we will download, and ingest.

An initial example will demonstrate that {duckdb) can invoke actual SQL commands. Even though we'll prefer to use {dplyr} functions most of the time, it's good to know that highly precise SQL commands can be invoked when needed.

```{r}
library(conflicted)
library(duckplyr)
library(fs)
library(nycflights13)

conflict_prefer("filter", "dplyr", quiet = TRUE)
```

The data and example code are found in [a DuckDB blog post](https://duckdb.org/2024/10/09/analyzing-open-government-data-with-duckplyr.html).

NOTE: The zipped data are \~ 100 MB. If you're on a slow data connection, you may find this download is slow.

```{r}
fs::dir_create("data")
download.file("https://blobs.duckdb.org/nzcensus.zip", "data/nzcensus.zip")
unzip("data/nzcensus.zip", exdir = "data")
```

## File system

Let's have a quick look at the unzipped files.

```{r}
fs::file_info(fs::dir_ls("data", glob = "*.csv"))
```

Take a quick look at the first 10 rows of CSV data.

```{r}
cat(paste(readLines("data/Data8277.csv", n=10), collapse="\n"))
```

## Select SQL examples

If you know SQL, you can still use without leveraging {dbplyr} as the SQL broker between {dplyr} verbs and the database engine. Let's have a quick look at some raw SQL, and then move back to {duckplyr}.

```{r}
duckdb:::sql("SELECT version()")
```

```{r}
duckdb:::sql("FROM 'data/Data8277.csv' LIMIT 10")
duckdb:::sql("FROM 'data/DimenLookupAge8277.csv' LIMIT 10")
```

```{r}
duckdb:::sql("DESCRIBE FROM 'data/Data8277.csv'")
duckdb:::sql("DESCRIBE FROM 'data/DimenLookupAge8277.csv'")
```

```{r}
duckdb:::sql("SUMMARIZE FROM 'data/Data8277.csv'")
```

## Duckplyr `read_csv_duckdb()`

Below we use the `read_csv_duckdb()`{=} function of {duckplyr} to import CSV files as DuckDB objects.

```{r}
data <- duckplyr::read_csv_duckdb("data/Data8277.csv")
data |> head()


age <- duckplyr::read_csv_duckdb("data/DimenLookupAge8277.csv")
area <- duckplyr::read_csv_duckdb("data/DimenLookupArea8277.csv")
ethnic <- duckplyr::read_csv_duckdb("data/DimenLookupEthnic8277.csv")
sex <- duckplyr::read_csv_duckdb("data/DimenLookupSex8277.csv"   )
year <- duckplyr::read_csv_duckdb("data/DimenLookupYear8277.csv"  )
# rm(data, age, area, ethnic, sex, year)
```

```{r}
class(data)
data |> explain()
```

```{r}
data |> 
  count() |> 
  mutate(count_pretty = scales::comma(n))
```

```{r}
data |> 
  head()
```

And here are the other objects imported from CSV files.

```{r}
age |> head()
area  |> head()
ethnic  |> head()
sex  |> head()
year  |> head()
```

### dplyr verbs for join

Leveraging {duckplyr} functions by using the verbs we know from {dplyr}.

```{r}
data  <- data |> 
  left_join(age, by = join_by("Age" == "Code")) |> 
  left_join(area, by = join_by("Area" == "Code")) |> 
  left_join(ethnic, by = join_by("Ethnic" == "Code")) |> 
  left_join(sex, by = join_by("Sex" == "Code")) |> 
  left_join(year, by = join_by("Year" == "Code"))
```

The output from this next code chunk is a lot of technical explanation.

```{r}
data |> 
  explain()
```

Instead of `left_join()`{=} we can also use `inner_join()`{=icml}, along with many other {dplyr} data wrangling functions.

```{r}
expanded_cleaned_data <- data |> 
  filter(grepl("^\\d+$", count)) |>
  mutate(count_ = as.integer(count)) |>
  filter(count_ > 0) |>
  inner_join(
    age |>
      filter(grepl("^\\d+ years$", Description)) |>
      mutate(age_ = as.integer(Code)),
    join_by(Age == Code)
  ) |> 
  inner_join(area |>
    mutate(area_ = Description) |>
    filter(!grepl("^Total", area_)), join_by(Area == Code)) |>
  inner_join(ethnic |>
    mutate(ethnic_ = Description) |>
    filter(!grepl("^Total", ethnic_)), join_by(Ethnic == Code)) |>
  inner_join(sex |>
    mutate(sex_ = Description) |>
    filter(!grepl("^Total", sex_)), join_by(Sex == Code)) |>
  inner_join(year |> mutate(year_ = Description), join_by(Year == Code))
```

### Collect()

Depending on your RAM, the `collect()`{=} function might now work. If not, `head()`{=} will be a quick-and-dirty way to get some output. Uncomment the line below; see what happens.

```{r}
expanded_cleaned_data |> 
  # collect()
  head()
```

### Lavish v stingy

Read more about the [levels of prudence](https://duckplyr.tidyverse.org/articles/prudence.html#concept). Which line, below, materializes the data into a *lavish* computation?

```{r}
# lavish v stingy
twenty_till_fourty_non_european_in_auckland_area <-
  expanded_cleaned_data |>
  duckplyr::as_duckdb_tibble(prudence = "lavish") |>
  filter(
    age_ >= 20, age_ <= 40,
    grepl("^Auckland", area_),
    year_ == "2018",
    ethnic_ != "European"
  ) |>
  summarise(group_count = sum(count_), .by = sex_) |> arrange(sex_)

twenty_till_fourty_non_european_in_auckland_area |> 
  mutate(count_pretty = scales::comma(group_count))
```
